---
title: "SleepAloneProject"
author: "Vee Kalkunte"
date: "February 18, 2020"
output: html_document
---

#Sleeping Alone Project: An Investigation Into Private Lives
##By Prithvi, "Vee" Kalkunte

Welcome to the Sleeping alone project: this project investigates trends in why couples/singles sleep alone rather than with another human being.

##Our cleaned up dataset

```{r cleaning}
d <- data[-c(1), -c(4, 5, 6, 7, 18, 19, 20, 21, 22, 23, 24)]
names(data) <- c("relationship", "length", "sleepTogetherFrequency", 
                 "remove", "remove", "remove", "remove", "snores", "bathroomTrips", "sick", "noSex", 
                 "temperature", "argument", "noSpace", "blanketHog", "child", "workHours", "remove",
                 "remove", "stayTogether", "sleepBetter", "sexLife", "occupation", "occupationOther",
                 "gender", "ageGroup", "income", "education", "cencusLocation")

#turn the checkboxes into booleans
d <- d %>% mutate(snores = ifelse(is.na(snores), FALSE, TRUE))
d <- d %>% mutate(bathroomTrips = ifelse(is.na(bathroomTrips), FALSE, TRUE))
d <- d %>% mutate(sick = ifelse(is.na(sick), FALSE, TRUE))
d <- d %>% mutate(noSex = ifelse(is.na(noSex), FALSE, TRUE))
d <- d %>% mutate(temperature = ifelse(is.na(temperature), FALSE, TRUE))
d <- d %>% mutate(argument = ifelse(is.na(argument), FALSE, TRUE))
d <- d %>% mutate(noSpace = ifelse(is.na(noSpace), FALSE, TRUE))
d <- d %>% mutate(blanketHog = ifelse(is.na(blanketHog), FALSE, TRUE))
d <- d %>% mutate(child = ifelse(is.na(child), FALSE, TRUE))
d <- d %>% mutate(workHours = ifelse(is.na(workHours), FALSE, TRUE))

#turn respective character columns to factor columns (they were multiple choice originally)
d$relationship <- factor(d$relationship, levels = c("Single", "Single, but cohabiting with a significant other",
                                                       "In a domestic partnership or civil union", "Married",
                                                       "Seperated", "Divorced", "Widowed"))
d$length <- factor(d$length, levels = c("Less than 1 year", "1-5 years", "6-10 years", "11-15 years",
                                           "16-20 years", "More than 20 years"))
d$sleepTogetherFrequency <- factor(d$sleepTogetherFrequency, levels = c("Never", "Once a year or less", "Once a month or less", "A few times per month",
                                                                           "A few times per week", "Every night"))
d$gender <- as.factor(d$gender)
d$ageGroup <- factor(d$ageGroup, levels = c("18-29", "30-44", "45-60", "> 60"))
d$income <- factor(d$income, levels = c("$0 - $24,999", "$25,000 - $49,999", "$50,000 - $99,999", "$100,000 - $149,999", "$150,000+"))
d$education <- factor(d$education, levels = c("Less than high school degree", "High school degree", 
                                              "Some college or Associate degree", "Bachelor degree", "Graduate degree"))
d$cencusLocation <- as.factor(d$cencusLocation)

#add in columns for number of reasons
d$numReasons <- rowSums(d == TRUE, na.rm = TRUE)
d <- d %>% mutate(hasReason = ifelse(numReasons > 0, TRUE, FALSE)) #1 is true
withReason <- d[d[, "numReasons"] > 0, ] #this set ticked off one of the reasons listed above
withoutReason <-d[d[, "numReasons"] <= 0, ] #the dataset that ticked off,,, nothing?

##what's the most common reason in the OneReason catagory?
oneReason <- d[d[, "numReasons"] == 1, ]

#need to mutate a new column with if elses that finds the one and puts it in a factored column
oneReason <- oneReason %>% mutate(reason = ifelse(snores, "snores", 
                                     ifelse(bathroomTrips, "bathroom", 
                                     ifelse(sick, "sick",
                                     ifelse(noSex, "noSex",
                                     ifelse(temperature, "temperature",
                                     ifelse(argument, "argument",
                                     ifelse(noSpace, "noSpace",
                                     ifelse(blanketHog, "blanketHog",
                                     ifelse(child, "child",
                                     ifelse(workHours, "work",NA)))))))))))
oneReason$reason <- as.factor(oneReason$reason)


head(d)
```

So a couple observations: our dataset is large, with a little over 1000 respondents, not all of them in relationship and not all of them not sleeping with their partners. To gain a better understanding of some of the demographic information of our dataset, let's throw in some ggplots.

##What does our dataset look like?

```{r plot1}
d %>% filter(!is.na(education)) %>% ggplot(aes(education)) + geom_bar() + coord_flip()
```

From the above, we can see that our sample is highly educated: we have a disproportionate representation of highly educated people in our sample. Now, with the following observations, we can see why this may be particularly heightened: 


```{r plot1}
d %>% filter(!is.na(ageGroup)) %>% ggplot(aes(x = ageGroup)) + geom_bar()
```

Our observation is older (and thus had more time to gain education). Furthermore, 

```{r plot1}
d %>% filter(!is.na(income)) %>% ggplot(aes(x = income)) + geom_bar()
```

Our observation is wealthier - all of which put the education into frame.

##How often do people sleep together? Are there any general trends in frequency?

###Observing the frequency of sleeping together based on education:
```{r}
withReason %>% filter(!is.na(sleepTogetherFrequency) & !is.na(education) & education != "Less than high school degree") %>% 
  ggplot(aes(education, fill = sleepTogetherFrequency)) + ylab("Percent") +
  geom_bar(position = "fill") + coord_flip() + scale_fill_manual(values=c("#540D6E", 
                                                                          "#EE4266", "#FFD23F", "#3BCEAC", "#06823C"))
```

We see that theres an observable negative trend between education and whether people sleep together.

###Here's the association between wealth and sleeping together frequency
```{r}
withReason %>% filter(!is.na(sleepTogetherFrequency) & !is.na(income)) %>% 
  ggplot(aes(income, fill = sleepTogetherFrequency)) + ylab("Percent") +
  geom_bar(position = "fill") + coord_flip() + scale_fill_manual(values=c("#540D6E", 
                                                                          "#EE4266", "#FFD23F", "#3BCEAC", "#06823C"))

```

###Here's a trend that shows length of relationship and sleeping together frequency
```{r}
withReason %>% filter(!is.na(sleepTogetherFrequency)) %>% 
  ggplot(aes(length, fill = sleepTogetherFrequency)) + ylab("Percent") +
  geom_bar(position = "fill") + coord_flip() + scale_fill_manual(values=c("#540D6E", 
                                    "#EE4266", "#FFD23F", "#3BCEAC", "#06823C"))

```

Something particularly interesting to note is that *both* people who sleep together every night and people who never sleep together generally increase with the length of time people stay together.

##Why do people sleep seperately?

First, we need to find the dataset of people who give one reason (becuase we don't have an indicator of the importance of each reason, and so multi reasoned individuals are not particularly beneficial), and then create the reasons variable, which gives the major reason as a factor we can put into a barplot.

```{r}
#add in columns for number of reasons
d$numReasons <- rowSums(d == TRUE, na.rm = TRUE)
d <- d %>% mutate(hasReason = ifelse(numReasons > 0, TRUE, FALSE)) #1 is true
withReason <- d[d[, "numReasons"] > 0, ] #this set ticked off one of the reasons listed above
withoutReason <-d[d[, "numReasons"] <= 0, ] #the dataset that ticked off nothing

##what's the most common reason in the OneReason catagory?
oneReason <- d[d[, "numReasons"] == 1, ]

#need to mutate a new column with if elses that finds the one and puts it in a factored column
oneReason <- oneReason %>% mutate(reason = ifelse(snores, "snores", 
                                     ifelse(bathroomTrips, "bathroom", 
                                     ifelse(sick, "sick",
                                     ifelse(noSex, "noSex",
                                     ifelse(temperature, "temperature",
                                     ifelse(argument, "argument",
                                     ifelse(noSpace, "noSpace",
                                     ifelse(blanketHog, "blanketHog",
                                     ifelse(child, "child",
                                     ifelse(workHours, "work",NA)))))))))))
oneReason$reason <- as.factor(oneReason$reason)
```

```{R}
#most common reasons
oneReason %>% filter(!is.na(reason)) %>% group_by(reason) %>% ggplot(aes(x = reason)) + geom_bar()
```

##Are there major reason differences between subsections?

###Let's see if there are reason differences between the sexes:
```{r}
oneReason %>% filter(!is.na(reason) & !is.na(gender)) %>% group_by(reason) %>% ggplot(aes(x = reason, fill = gender)) + geom_bar(position = "fill") + ylab("Percent") 

```
It looks like the most "male dominated" complaints are temperature differences and bathroom visits, while the most "female dominated" complaints are snoring and arguments. However, taking the ggplot out of fill, we see something that may contest that:


```{r}
oneReason %>% filter(!is.na(reason) & !is.na(gender)) %>% group_by(reason) %>% ggplot(aes(x = reason, fill = gender)) + geom_bar()

```

For the most part, the larger the sample size there is, the closer it is to an even split (with the only exception being the reason that one of the spouses needed to sleep with a child). And so, there's reason to believe that the domination isn't necessarily reliable.

###To investigate this stration by income:
```{r}
oneReason %>% filter(!is.na(reason) & !is.na(income)) %>% group_by(income) %>% ggplot(aes(x = income, fill = reason)) + geom_bar(position = "fill") + ylab("Percent") + coord_flip()

oneReason %>% filter(!is.na(reason) & !is.na(income)) %>% group_by(income) %>% ggplot(aes(x = income, fill = reason)) + geom_bar() + coord_flip()


```

Switching between the fill and unfilled barplots, we see a few interesting notes:

- Can't make major conclusions with the "0-24,999" catagory becuase of low count; but if we could, sickness is the most common reason within the sample
- More or less similar across most groups, though snoring is the biggest reason within the "100,000-149,999" catagory.
- An interesting note: work reasons stay even across economic catagories.

Income is likely not a good reason to predict the reason why people don't sleep together, becuase, although there are minor trends, for the most part the splits are the same.
 
###To investigate this stration by education:
```{r}
oneReason %>% filter(!is.na(reason) & !is.na(education)) %>% group_by(education) %>% ggplot(aes(x = education, fill = reason)) + geom_bar(position = "fill") + ylab("Percent") + coord_flip()

oneReason %>% filter(!is.na(reason) & !is.na(education)) %>% group_by(education) %>% ggplot(aes(x = education, fill = reason)) + geom_bar() + coord_flip()

```

###To investigate by agegroup:
```{r}
oneReason %>% filter(!is.na(reason) & !is.na(ageGroup)) %>% group_by(ageGroup) %>% ggplot(aes(x = ageGroup, fill = reason)) + geom_bar(position = "fill") + ylab("Percent") + coord_flip()

oneReason %>% filter(!is.na(reason) & !is.na(ageGroup)) %>% group_by(ageGroup) %>% ggplot(aes(x = ageGroup, fill = reason)) + geom_bar() + coord_flip()
```

Some interesting notes:

- The issue of work differences and argument differences  have a negative correlation with age
- Snoring seems to be the biggest problem in the "45-60" catagory, but after 60, sickness takes the lead (though snoring still stays a major factor)
- child reasons are mainly within the "30-44" catagory

###To investigate by length of relationship:
```{r}
oneReason %>% filter(!is.na(reason) & !is.na(length)) %>% group_by(length) %>% ggplot(aes(x = length, fill = reason)) + geom_bar(position = "fill") + ylab("Percent") + coord_flip()

oneReason %>% filter(!is.na(reason) & !is.na(length)) %>% group_by(length) %>% ggplot(aes(x = length, fill = reason)) + geom_bar() + coord_flip()
```

Some notes:
- Discounting "less than one year" becuase of the small amount
- Reasons seem to be pretty stable across length, though the oldest relationshups show snoring and illness as the major reasons, similar to age groups
- Arguments tend to be a bigger reason when the relationship is newer, but as it gets older, it's no longer a particularly large reason.

##Conclusions: What did we learn today?

From our dataset, biased towards a highly educated, wealthy, and older demographic with longer relationships, we have found the following conclusions:

###Who sleeps together?

Most frequently? Older relationships, couples within the "25,000 - 49,999" income bracket.

###Who doesn't sleep together?

Younger couples, or richer couples.

###Why don't people sleep together?

Most commonly, sickness and snoring, with work, arguments and children trailing after - though, of course, certain subsections show different proportions of each, with age and length of relationship being the most telling factors in determining the reason.


